---
title: "Valtýr-GPT"
author: "Valtýr"
date: "2023-04-26"
output: html_document
---

```{r pakkar, message=FALSE, warning=FALSE, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
#Rpakkar <- c("tidyverse","benthos","worrms", "plyr", "data.table", "DT", "dplyr", "here", "BBI", "ggh4x", "gplots","PDE", "magick", "tesseract", "xlsx", "ggpubr", "factoextra")
Rpakkar <- c("tidyverse", "diagram", "genogram", "xlsx", "vegan", "plyr", "ggrepel")

#install.packages('pacman')
pacman::p_load(Rpakkar, character.only = TRUE)
#remotes::install_github(c("ropensci/tabulizerjars", "ropensci/tabulizer"))
#install.packages("PDE", dependencies = TRUE)
#Sys.setlocale("LC_ALL", "Icelandic") # `Sys.setlocale` er hér fyrir íslenskar dagsetningar.
```

```{r innlestur}
KolgrTaxa <- read_csv("KolgrTaxa.csv", na = "empty") 
   
   remove_list <- paste(c(
     "nýsestir",
     "ungviði",
     "ungv",
     "ungv.",
     "juv",
     "harpacticoida",
     "Acarina sp"
   ), collapse = '|') 
   
   remove_ind <- lapply(strsplit(remove_list , "\\|")[[1]] , \(x) grep(x , KolgrTaxa$gamalt , fixed = T)) |> 
     unlist() |> 
     unique()
   
   ekkiungvidi <- KolgrTaxa[-remove_ind,] 
   
   df <- ekkiungvidi %>% 
     mutate(Artal = factor(Artal)) %>% 
     filter(stod %in% c("C4", "A7", "B5", "B8", "E4", "E3")) %>% 
     ddply(.(Artal, stod,Flokkun),summarise, N=sum(N)) %>% 
     arrange(N)
   
   jorundur <- df %>% 
     filter(!Flokkun %in% c("harpacticoida", "Campanulariidae")) %>% 
     mutate(
       Flokkun = case_when(
         Flokkun == "Spionida" ~ "Spionidae",
         Flokkun == "Terebellides stroemi" ~ "Terebellides stroemii",
         Flokkun == "Ampharetinae" ~ "Ampharetidae",
         Flokkun == "ostracoda" ~ "Ostracoda",
         Flokkun == "Sabellidae" ~ "Sabellida",
         Flokkun == "scalibregma inflatum" ~ "Scalibregma inflatum",
         Flokkun == "sipunculida" ~ "Sipuncula",
         Flokkun == "sipunculidae" ~ "Sipuncula",
         Flokkun == "sipunculidea" ~ "Sipuncula",
         Flokkun == "Nephtys" ~ "Nephthys",
         Flokkun == "Nepthys" ~ "Nephthys",
Artal == "1999" & Flokkun == "Ampharete acutifrons" ~ "Ampharetinae",
Artal == "1999" & Flokkun == "Bivalvia" ~ "NA"        ,
Artal == "2013" & Flokkun == "Harmothoe" ~ "Harmothoe extenuata"       ,
Artal == "2013" & Flokkun == "Polynoidae" ~ "Harmothoe extenuata" ,
Artal == "2014" & Flokkun == "Praxillella" ~ "Praxillella praetermissa",
Artal == "2014" & Flokkun == "Syllidae" ~ "Syllis cornuta"             ,
Artal == "2015" & Flokkun == "Syllidae" ~ "Syllis cornuta",
Artal == "2015" & Flokkun == "Mya" ~ "Mya arenaria"        ,     
Artal == "2015" & Flokkun == "Mytilidae" ~ "Mytilus edulis",
Artal == "2016" & Flokkun == "Aricidea" ~ "Amphitrite cirrata"      ,
Artal == "2016" & Flokkun == "Capitellidae" ~ "Capitella capitata"   ,  
Artal == "2016" & Flokkun == "Cirratulidae" ~ "Cirratulus cirratus",
Artal == "2016" & Flokkun == "Cossuridae" ~ "Cossura longocirrata",
Artal == "2016" & Flokkun == "Nephtyidae" ~ "Nephthys"      ,
Artal == "2016" & Flokkun == "Pectinariidae" ~ "Pectinaria koreni"     ,
Artal == "2016" & Flokkun == "Phyllodocida" ~ "Phyllodoce maculata",
Artal == "2016" & Flokkun == "Spio" ~ "Spio filicornis"            ,
Artal == "2016" & Flokkun == "Spionidae" ~ "Spio filicornis",
Artal == "2016" & Flokkun == "Syllidae" ~ "Syllis"           ,          
Artal == "2016" & Flokkun == "Cardiidae" ~ "Cardium"          ,         
Artal == "2016" & Flokkun == "Cardiidae" ~ "Cardium"           ,        
Artal == "2017" & Flokkun == "Mya" ~ "Mya arenaria"  ,
Artal == "2017" & Flokkun == "Maldanidae" ~ "Praxillella praetermissa",
Artal == "1999" & Flokkun == "Ampharetinae" ~ "Ampharetidae",
Artal == "2017" & Flokkun == "Ampharete" ~ "Ampharetidae"    ,
Artal == "2017" & Flokkun == "Ampharete acutifrons" ~ "Ampharetidae",
Artal == "2016" & Flokkun == "Lumbrineridae" ~ "Lumbrineris",
Artal == "2015" & Flokkun == "Pholoe" ~ "Pholoe minuta"      ,          
Artal == "2016" & Flokkun == "Pholoe" ~ "Pholoe minuta"       ,         
Artal == "2017" & Flokkun == "Pholoe" ~ "Pholoe minuta"        ,    
Artal == "1999" & Flokkun == "Musculus discors" ~ "Musculus",
Artal == "2014" & Flokkun == "Mya" ~ "Mya arenaria"          ,   
Artal == "2014" & Flokkun == "Mytilus edulis" ~ "Mytilidae",
Artal == "2014" & Flokkun == "Nephtyidae" ~ "Nephthys"       ,
Artal == "2014" & Flokkun == "Amphipoda" ~ "Protomedeia fasciata"    ,
Artal == "2014" & Flokkun == "Tubificidae" ~ "Tubificoides kozloffi",
Artal == "2014" & Flokkun == "Harmothoe" ~ "Harmothoe extenuata"     ,  
Artal == "2015" & Flokkun == "Harmothoe" ~ "Harmothoe extenuata"      , 
Artal == "2015" & Flokkun == "Amphipoda" ~ "Protomedeia fasciata",
Artal == "2016" & Flokkun == "Balanus balanus" ~ "Balanus"       ,
Artal == "2016" & Flokkun == "priapulidae" ~ "Priapulus caudatus" ,     
Artal == "2016" & Flokkun == "Priapulidae" ~ "Priapulus caudatus"  ,    
Artal == "2017" & Flokkun == "Harmothoe" ~ "Harmothoe extenuata"    ,   
Artal == "1999" & Flokkun == "Leucon acutirostris" ~ "Cumacea"       ,  
Artal == "1999" & Flokkun == "Eudorella emarginata" ~ "Cumacea",
       TRUE ~ Flokkun)) %>% 
     drop_na() 
```


```{r environmental}
kornast_organic_tap_dypi <- structure(list(stod = c("A7", "A7", "A7", "A7", "A7", "B5", "B5", 
"B5", "B5", "B5", "B8", "B8", "B8", "B8", "B8", "C4", "C4", "C4", 
"C4", "C4", "E3", "E3", "E3", "E3", "E3", "E4", "E4", "E4", "E4", 
"E4"), Artal = c("2013", "2014", "2015", "2016", "2017", "2013", 
"2014", "2015", "2016", "2017", "2013", "2014", "2015", "2016", 
"2017", "2013", "2014", "2015", "2016", "2017", "2013", "2014", 
"2015", "2016", "2017", "2013", "2014", "2015", "2016", "2017"
), `20um` = c(0.628496042216359, 0.343339587242026, 0.453474676089517, 
0.390734265734266, 0.643465909090909, 0.386002120890774, 0.321766561514196, 
0.386792452830189, 0.19559585492228, 0.546637744034707, 0.592529025744573, 
0.515981735159817, 0.369913686806412, 0.424997544719429, 0.221565731166913, 
0.332928794917986, 0.153116531165312, 0.234791889007471, 0.190773067331671, 
0.152190051967335, 0.385025817555938, 0.211453744493392, 0.157303370786517, 
0.0827048768225239, 0.247269116186693, 0.426682936750051, 0.406354515050167, 
0.214331413947537, 0.0125582337451894, 0.286830357142857), `63um` = c(0.350395778364116, 
0.24577861163227, 0.215547703180212, 0.215909090909091, 0.151988636363636, 
0.52661717921527, 0.342271293375394, 0.304245283018868, 0.348445595854922, 
0.31236442516269, 0.317491166077739, 0.23972602739726, 0.276202219482121, 
0.278886610994081, 0.282127031019202, 0.306115483075756, 0.276422764227642, 
0.257203842049093, 0.17643391521197, 0.197475872308834, 0.379173838209983, 
0.177679882525698, 0.143392188336009, 0.0935143288084465, 0.200595829195631, 
0.438071995118975, 0.29933110367893, 0.211772232885477, 0.0256734859226251, 
0.266741071428571), `125um` = c(0.0100263852242744, 0.315196998123827, 
0.267373380447585, 0.284965034965035, 0.177556818181818, 0.0752916224814422, 
0.32807570977918, 0.295990566037736, 0.399611398963731, 0.134490238611714, 
0.0261736496718829, 0.237442922374429, 0.327990135635019, 0.240590007791825, 
0.370753323485968, 0.319370437091116, 0.485094850948509, 0.44076840981857, 
0.350997506234414, 0.528582034149963, 0.207194492254733, 0.281938325991189, 
0.173354735152488, 0.283810960281549, 0.201588877855015, 0.104535285743339, 
0.249163879598662, 0.19065898912348, 0.0339781243670245, 0.280133928571429
), `250um` = c(0.0058047493403694, 0.0956848030018762, 0.0459363957597173, 
0.0734265734265734, 0.00852272727272727, 0.00742311770943797, 
0.00788643533123028, 0.0129716981132075, 0.0518134715025907, 
0.00650759219088937, 0.0553003533568905, 0.00684931506849315, 
0.0135635018495684, 0.0384999321551487, 0.0782865583456425, 0.0337347112923107, 
0.0569105691056911, 0.0448239060832444, 0.180174563591022, 0.0972531551596139, 
0.0236660929432013, 0.0558002936857562, 0.0358480470840021, 0.0834590246354952, 
0.041708043694141, 0.0213544844417328, 0.0133779264214047, 0.0300703774792067, 
0.00212679765039498, 0.0223214285714286), `1000um` = c(0.00527704485488127, 
0, 0.0176678445229682, 0.034965034965035, 0.0184659090909091, 
0.00466595970307529, 0, 0, 0.00453367875647668, 0, 0.00850580514891469, 
0, 0.0123304562268804, 0.0170259043395175, 0.0472673559822747, 
0.00785057362283114, 0.0284552845528455, 0.0224119530416222, 
0.101620947630923, 0.0244988864142539, 0.00493975903614458, 0.273127753303965, 
0.490101658640984, 0.456510809451986, 0.30883813306852, 0.00935529794590197, 
0.0317725752508361, 0.353166986564299, 0.925663358314766, 0.143973214285714
), tap = c(0.181515414138913, 0.117573294968505, 0.118865354458575, 
0.10855175801997, 0.144215480661367, 0.174146595044186, 0.119127447894571, 
0.124407020872865, 0.105198658268468, 0.115848718019756, 0.178503108560497, 
0.12940849965294, 0.117179453375411, 0.111245191819955, 0.116370341491164, 
0.142787808192584, 0.113006923837784, 0.132290387475618, 0.0969106601341345, 
0.10220582768636, 0.215412621359223, 0.101006035443262, 0.090402518790167, 
0.0885715142546809, 0.0796504565039956, 0.135196812364163, 0.126232974506719, 
0.103923910552975, 0.092065868263473, 0.101525813669742), sd = c(0.00544534378975074, 
0.000976059208807937, 0.00105675176697187, 0.00129099190584701, 
0.010280862872346, 0.00245055200549895, 0.000348361831630225, 
0.00083859912379809, 0.001823914492673, 0.000495255850064288, 
0.010990421037094, 0.000600243017338078, 0.00277626938387639, 
0.000420021748488332, 0.0014744503375705, 0.0199049666612412, 
0.00034970661779751, 0.00314117917812511, 0.000215745910008022, 
0.000435189983246506, 0.127862755335917, 0.000119010118739825, 
0.00226462935582809, 0.00132600786385895, 0.00243540659124004, 
0.0118334943096431, 0.000449569977393725, 0.000285425329099763, 
0.00246993386642007, 0.000814994916882806), dypi = c(16.6, 16.6, 
16.6, 16.6, 16.6, 24.5, 24.5, 24.5, 24.5, 24.5, 16.6, 16.6, 16.6, 
16.6, 16.6, 34, 34, 34, 34, 34, 11.2, 11.2, 11.2, 11.2, 11.2, 
11, 11, 11, 11, 11), `Síld 2013` = c(0L, 0L, 0L, 0L, 0L, 1L, 
1L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L)), row.names = c(NA, -30L), class = "data.frame")

```


```{r MergeSep}

# Aggregate data to sum 'N' values for duplicate rows
#data1_agg <- jorundur %>% group_by(Artal, stod, Flokkun) %>% summarise(N = sum(N))
data1_agg <- ddply(jorundur,.(Artal, stod, Flokkun),summarise,N=sum(N))
# Prepare data for ordination
data1_wide <- data1_agg %>% spread(key = Flokkun, value = N, fill = 0)

merged_data <- merge(data1_wide, kornast_organic_tap_dypi, by = c("Artal", "stod")) # Hér dettur 1999 út
rownames(merged_data) <- paste(merged_data$Artal, merged_data$stod, sep = "_")
merged_data <- merged_data[, -c(1,2)] # Remove the 'Year' and 'Station' columns

# Separate environmental and species data
env_data <- merged_data[, c("20um","63um","125um","250um","1000um","tap","sd","dypi", "Síld 2013")]
species_data <- merged_data[, !(names(merged_data) %in% c("20um","63um","125um","250um","1000um","tap","sd","dypi", "Síld 2013"))]

```



```{r Decorana, message=FALSE, warning=FALSE}
dca <- decorana(species_data)
dca # fyrsti DCA-ásinn eftir decorana() er >4
# Get sample scores
sample_scores <- scores(dca, display = "sites")
species_scores <- scores(dca, display = "species")
# Combine sample and species scores into a single data frame
combined_scores <- rbind(data.frame(Scores = sample_scores, Labels = row.names(sample_scores), Type = "Sample"),data.frame(Scores = species_scores, Labels = row.names(species_scores), Type = "Species"))


# Create the DCA plot með ggplot
p <- ggplot(combined_scores, aes(x = Scores.DCA1, y = Scores.DCA2, color = Type, label = Labels)) +
    geom_point() +
    geom_text_repel(size = 3, max.overlaps = Inf) + # Increase max.overlaps to allow more overlaps
    labs(title = "Detrended Correspondence Analysis (DCA)",
         x = "DCA Axis 1",
         y = "DCA Axis 2") +
    theme_minimal() +
    scale_color_manual(values = c("Sample" = "black", "Species" = "red"))

  plotly::ggplotly(p)
  #p
```

Based on the DCA output provided, the gradient lengths for the first and second axes are 4.0100 and 2.5018, respectively.

The first DCA axis has a gradient length greater than 4, which indicates a strong species turnover along the primary gradient. This suggests that unimodal ordination methods, such as CA or CCA, are more appropriate for the data, as these methods are designed to handle data with strong species turnover.

The second DCA axis has a gradient length of 2.5018, which indicates a weaker species turnover along the secondary gradient. In this case, linear ordination methods like PCA or RDA might be suitable for analyzing patterns along the secondary gradient. However, since the primary gradient is more critical in determining the overall pattern of species composition, it's usually recommended to choose an ordination method based on the primary gradient.

In summary, given the gradient lengths from your DCA output, it is advisable to use unimodal ordination methods like CA or CCA for your analysis, as they are more suitable for handling the strong species turnover observed along the primary gradient in your data.


*Ég*: what is a species turnover along the primary gradient?
*ChatGPT*
Species turnover, also known as beta diversity, refers to the change in species composition across different sites or samples along a gradient. The primary gradient typically represents the main underlying environmental factor or combination of factors that drive the observed changes in species composition in the dataset.

In the context of ordination methods, a species turnover along the primary gradient means that there is a substantial change in the species composition from one end of the gradient to the other. This turnover can be due to various reasons, such as differing environmental conditions, resource availability, or competition among species.

When the species turnover along the primary gradient is strong, unimodal ordination methods like Correspondence Analysis (CA) or Canonical Correspondence Analysis (CCA) are more suitable, as they are designed to handle data with a high degree of species turnover. These methods are based on the assumption that species show a unimodal response to underlying environmental gradients, meaning that each species has an optimal range of conditions along the gradient where it is most abundant. Unimodal ordination methods can better capture the patterns of species composition when species exhibit strong turnover along the primary gradient.

In contrast, when species turnover along the primary gradient is weak, linear ordination methods like Principal Components Analysis (PCA) or Redundancy Analysis (RDA) may be more appropriate. These methods assume that species exhibit linear relationships with environmental variables, and they are better suited for datasets with less pronounced species turnover along the gradients.



#CCA
Given the gradient lengths from your DCA output, the next step is to perform a unimodal ordination method, such as Canonical Correspondence Analysis (CCA), to analyze the relationship between species composition and environmental variables.
```{r CCA}

cca_result <- cca(species_data ~ ., data = env_data)
plot(cca_result)

# Add points with the labels
text(cca_result, display = "sites", col = "blue")

anova_result <- anova(cca_result, by = "terms", perm = 999)
print(anova_result)


```



The output you provided shows the results of the permutation test for the Canonical Correspondence Analysis (CCA) under the reduced model. The test assesses the significance of the environmental variables in explaining the species composition in your data.

The table displays the degrees of freedom (Df), Chi-square values, F statistics, and p-values (Pr(>F)) for each environmental variable, as well as the residual. The p-values indicate the significance of the relationship between each environmental variable and species composition, with smaller p-values suggesting a stronger relationship.

According to the results:

20um (***): Highly significant relationship (p < 0.001)
63um (**): Significant relationship (p = 0.007)
125um (*): Significant relationship (p = 0.033)
250um: Not significant (p = 0.240)
tap: Not significant (p = 0.182)
sd: Not significant (p = 0.730)
dypi: Not significant (p = 0.147)
Síld 2013: Not significant (p = 0.136)
Based on these results, it appears that the environmental variables 20um, 63um, and 125um have significant relationships with species composition in your data. The other environmental variables (250um, tap, sd, dypi, and Síld 2013) do not show significant relationships.

You can use these results to help interpret the CCA ordination plot, identify the main environmental drivers of species composition, and guide further analysis or data collection.



```{r CCA_B}

# Plot species points
plot(cca_result, display = "species", type = "n")
# Add species labels
text(cca_result, display = "species", cex = 0.8, col = "blue")

# Add environmental variable vectors
envfit_result <- envfit(cca_result, env_data)
plot(envfit_result, p.max = 0.1, col = "black") # Adjust p.max value to control the number of plotted vectors


```

fig. cap. Figure X: Canonical Correspondence Analysis (CCA) biplot displaying the relationships between species (blue text) and environmental variables (black arrows) in Kolgrafafjörður. Species points represent their position in the ordination space based on their abundance patterns across sampling stations, while the lengths and directions of the environmental variable vectors indicate the strength and direction of their relationships with the species composition. A shorter vector implies a weaker relationship, whereas a longer vector indicates a stronger association with the species distribution. The angles between the vectors represent the correlations between environmental variables, with acute angles indicating positive correlations and obtuse angles indicating negative correlations. Only environmental variables with p-values below 0.1 are displayed in the plot to focus on the most influential factors affecting the species composition.

# why do these results not agree?

The results from the envfit() function and the anova() function on your CCA model are addressing two different things, which is why they might not seem to agree.

The envfit() function in vegan performs a simple linear regression of the ordination scores (the positions of your sites or species in the ordination space) against each of your environmental variables separately. The reported r2 values are the coefficients of determination for these simple regressions, which measure the proportion of the variance in the ordination scores that is predictable from the environmental variable. The Pr(>r) values are the p-values for the significance of these regressions, tested by permutation.

The anova() function is performing a permutation test on the entire CCA model, testing the significance of each term after controlling for the effects of the others. This is a multivariate analysis, which is why the significance levels may differ from those in the univariate regressions performed by envfit(). The reported Pr(>F) values are the p-values for the significance of the improvement in the model when each term is added, tested by permutation.

So in essence, envfit() is looking at the relationships between the environmental variables and the ordination scores one at a time, whereas anova() is looking at the combined effects of all the environmental variables together. That's why you're seeing different results.

In your case, the environmental variables that are significant in the envfit() output are not necessarily significant in the anova() output. This can occur when there's multicollinearity among the environmental variables, i.e., when the variables are correlated with each other. In such cases, a variable that is significant on its own (according to envfit()) might not significantly improve the model when other variables are already included (according to anova()).

#NMDS
Do Non-metric Multidimensional Scaling (NMDS): Visualize community composition in a reduced-dimensionality space and overlay environmental variables to explore potential drivers of community structure.

To perform Non-metric Multidimensional Scaling (NMDS) and visualize community composition in a reduced-dimensionality space with environmental variables overlaid, follow these steps:

```{r NMDS}


#Calculate the dissimilarity matrix for species data using the Bray-Curtis distance:

species_dist <- vegdist(species_data, method = "bray")

#Perform NMDS:

nmds_result <- metaMDS(species_dist, k = 2)

#Visualize the NMDS ordination plot:

plot(nmds_result, type = "n", main = "NMDS Ordination")
text(nmds_result, display = "sites", col = "black")

#Fit environmental variables to the NMDS ordination:

envfit_result <- envfit(nmds_result, env_data)

#Overlay environmental variables on the NMDS plot:

plot(envfit_result, p.max = 0.05) # Only show significant variables at the 0.05 level

#These steps will help you perform NMDS and visualize the community composition in a two-dimensional space. Additionally, the significant environmental variables will be overlaid on the NMDS plot to explore their potential role in driving community structure. The resulting NMDS plot will provide a visual representation of the relationships between species, samples, and environmental variables in your data.

```
Figure X: Non-metric Multidimensional Scaling (NMDS) ordination plot illustrating the community composition of species in Kolgrafafjörður based on the Bray-Curtis dissimilarity matrix. Sampling stations are represented by black text, with their positions in the ordination space reflecting their relative similarity in species composition. The environmental variables significantly associated (p < 0.05) with the community structure are overlaid on the plot as black arrows, with the length and direction of the arrows indicating the strength and direction of their influence on the species composition. Acute angles between the environmental variable vectors represent positive correlations, while obtuse angles indicate negative correlations. The NMDS plot provides a visual representation of the relationships between species, samples, and environmental variables, aiding in the identification of potential drivers of community structure in the study area.

In an NMDS plot, the axes represent the reduced-dimensionality space that best preserves the original dissimilarities among samples in the high-dimensional species space. The axes themselves do not have a direct ecological interpretation, as they are derived from the ranked dissimilarities between samples. Instead, they serve as a means to visualize complex, multidimensional relationships among samples in a more interpretable two-dimensional space.

The NMDS plot helps to reveal patterns in community composition by positioning samples close to each other if they have similar species compositions and further apart if their compositions are dissimilar. In this way, the NMDS plot allows you to visually assess the relationships among samples and identify potential groupings or gradients in the data.

Overlaying environmental variables on the NMDS plot can help you explore their potential role in driving community structure. Arrows representing environmental variables are fitted to the ordination space, and the direction and length of these arrows provide information about the strength and direction of the relationships between these variables and the ordination axes. The angle between two arrows indicates the correlation between the corresponding environmental variables: a small angle represents a positive correlation, while an angle close to 180 degrees represents a negative correlation. If an environmental variable is significantly related to the community composition, its arrow will point in the direction of the gradient along which the variable changes most. By inspecting the NMDS plot with overlaid environmental variables, you can identify which factors might be important in shaping the observed community patterns.


### Feeding Guilds (*Sjá Shiny app*)
```{r feeding}
data1_agg <- ddply(jorundur,.(Artal, stod, Flokkun), summarise, N = sum(N))


#f <- file.path(list.files(here::here("set"),pattern = "pdf",full.names = T))
#tafla <- tabulizer::extract_tables(f, pages = 17:48)
#tabbla <- do.call(rbind,tafla)
tabbla_df <- as.data.frame(tabbla)
colnames(tabbla_df) <- c("Major Group", "Family code", "Species code", "Taxon name", "Food Source", 
                         "Motility", "Habit", "Om/Ca/He", "Food size/type", "FeedMode", 
                         "Feeding guild")
table_df <- subset(tabbla_df, !grepl("^Major", `Major Group`))


# nákvæm pörun:
find_exact_match <- function(x, y) {
  sapply(x, function(Flokkun) {
    matched <- y[Flokkun == y]
    if (length(matched) > 0) {
      return(matched[1])
    } else {
      return(NA_character_)
    }
  }, simplify = "character")
}


table_df$`Cleaned Taxon name` <- gsub("\\s*(sp\\.|spp\\.|indet\\.)", "", table_df$`Taxon name`)

data1_agg$exact_taxon_name <- find_exact_match(data1_agg$Flokkun, table_df$`Cleaned Taxon name`)
unmatched_rows <- is.na(data1_agg$exact_taxon_name)

 find_closest_match <- function(x, y) {
   string_distances <- stringdist::stringdistmatrix(tolower(x), tolower(y), method = "jw")
   y[apply(string_distances, 1, which.min)]
 }
  
data1_agg$matched_taxon_name[unmatched_rows] <- find_closest_match(data1_agg$Flokkun[unmatched_rows], table_df$`Cleaned Taxon name`)
data1_agg$final_taxon_name <- ifelse(is.na(data1_agg$exact_taxon_name), data1_agg$matched_taxon_name, data1_agg$exact_taxon_name)
merged_data <- merge(data1_agg, table_df, by.x = "final_taxon_name", by.y = "Taxon name")

# merged_data er 635 raðir á móti 816 röðum hjá data1_agg. Mikið sem dettur út.
```

```{r Feeding}

KolgrTaxa <- read_csv("KolgrTaxa.csv", na = "empty") 
   
   remove_list <- paste(c(
     "nýsestir",
     "ungviði",
     "ungv",
     "ungv.",
     "juv",
     "harpacticoida"
   ), collapse = '|') 
   
   remove_ind <- lapply(strsplit(remove_list , "\\|")[[1]] , \(x) grep(x , KolgrTaxa$gamalt , fixed = T)) |> 
     unlist() |> 
     unique()
   
   ekkiungvidi <- KolgrTaxa[-remove_ind,] 
   
   # df <- ekkiungvidi %>% 
   #   mutate(Artal = factor(Artal)) %>% 
   #   filter(stod %in% c("C4", "A7", "B5", "B8", "E4", "E3")) %>% 
   #   ddply(.(Artal, stod,Flokkun),summarise, N=sum(Nfm)) %>% 
   #   arrange(N)
   
   lagfaerd_heiti <- ekkiungvidi %>% 
     filter(!Flokkun %in% c("harpacticoida", "Campanulariidae")) %>% 
     mutate(
       Flokkun = case_when(
         Flokkun == "Spionida" ~ "Spionidae",
         Flokkun == "Terebellides stroemi" ~ "Terebellides stroemii",
         Flokkun == "Ampharetinae" ~ "Ampharetidae",
         Flokkun == "ostracoda" ~ "Ostracoda",
         Flokkun == "Sabellidae" ~ "Sabellida",
         Flokkun == "scalibregma inflatum" ~ "Scalibregma inflatum",
         Flokkun == "sipunculida" ~ "Sipuncula",
         Flokkun == "sipunculidae" ~ "Sipuncula",
         Flokkun == "sipunculidea" ~ "Sipuncula",
         Flokkun == "Nephtys" ~ "Nephthys",
         Flokkun == "Nepthys" ~ "Nephthys",
Artal == "1999" & Flokkun == "Ampharete acutifrons" ~ "Ampharetinae",
Artal == "1999" & Flokkun == "Bivalvia" ~ "NA"        ,
Artal == "2013" & Flokkun == "Harmothoe" ~ "Harmothoe extenuata"       ,
Artal == "2013" & Flokkun == "Polynoidae" ~ "Harmothoe extenuata" ,
Artal == "2014" & Flokkun == "Praxillella" ~ "Praxillella praetermissa",
Artal == "2014" & Flokkun == "Syllidae" ~ "Syllis cornuta"             ,
Artal == "2015" & Flokkun == "Syllidae" ~ "Syllis cornuta",
Artal == "2015" & Flokkun == "Mya" ~ "Mya arenaria"        ,     
Artal == "2015" & Flokkun == "Mytilidae" ~ "Mytilus edulis",
Artal == "2015" & Flokkun == "Ampharetidae" ~ "NA"          ,
Artal == "2016" & Flokkun == "Aricidea" ~ "Amphitrite cirrata"      ,
Artal == "2016" & Flokkun == "Capitellidae" ~ "Capitella capitata"   ,  
Artal == "2016" & Flokkun == "Cirratulidae" ~ "Cirratulus cirratus",
Artal == "2016" & Flokkun == "Cossuridae" ~ "Cossura longocirrata",
Artal == "2016" & Flokkun == "Nephtyidae" ~ "Nephthys"      ,
Artal == "2016" & Flokkun == "Pectinariidae" ~ "Pectinaria koreni"     ,
Artal == "2016" & Flokkun == "Phyllodocida" ~ "Phyllodoce maculata",
Artal == "2016" & Flokkun == "Spio" ~ "Spio filicornis"            ,
Artal == "2016" & Flokkun == "Spionidae" ~ "Spio filicornis",
Artal == "2016" & Flokkun == "Syllidae" ~ "Syllis"           ,          
Artal == "2016" & Flokkun == "Cardiidae" ~ "Cardium"          ,         
Artal == "2016" & Flokkun == "Cardiidae" ~ "Cardium"           ,        
Artal == "2017" & Flokkun == "Mya" ~ "Mya arenaria"  ,
Artal == "2017" & Flokkun == "Maldanidae" ~ "Praxillella praetermissa",
Artal == "1999" & Flokkun == "Ampharetinae" ~ "Ampharetidae",
Artal == "2017" & Flokkun == "Ampharete" ~ "Ampharetidae"    ,
Artal == "2017" & Flokkun == "Ampharete acutifrons" ~ "Ampharetidae",
Artal == "2016" & Flokkun == "Lumbrineridae" ~ "Lumbrineris",
Artal == "2015" & Flokkun == "Pholoe" ~ "Pholoe minuta"      ,          
Artal == "2016" & Flokkun == "Pholoe" ~ "Pholoe minuta"       ,         
Artal == "2017" & Flokkun == "Pholoe" ~ "Pholoe minuta"        ,    
Artal == "1999" & Flokkun == "Musculus discors" ~ "Musculus",
Artal == "2014" & Flokkun == "Mya" ~ "Mya arenaria"          ,   
Artal == "2014" & Flokkun == "Mytilus edulis" ~ "Mytilidae",
Artal == "2014" & Flokkun == "Nephtyidae" ~ "Nephthys"       ,
Artal == "2014" & Flokkun == "Amphipoda" ~ "Protomedeia fasciata"    ,
Artal == "2014" & Flokkun == "Tubificidae" ~ "Tubificoides kozloffi",
Artal == "2014" & Flokkun == "Harmothoe" ~ "Harmothoe extenuata"     ,  
Artal == "2015" & Flokkun == "Harmothoe" ~ "Harmothoe extenuata"      , 
Artal == "2015" & Flokkun == "Amphipoda" ~ "Protomedeia fasciata",
Artal == "2016" & Flokkun == "Balanus balanus" ~ "Balanus"       ,
Artal == "2016" & Flokkun == "priapulidae" ~ "Priapulus caudatus" ,     
Artal == "2016" & Flokkun == "Priapulidae" ~ "Priapulus caudatus"  ,    
Artal == "2017" & Flokkun == "Harmothoe" ~ "Harmothoe extenuata"    ,   
Artal == "1999" & Flokkun == "Leucon acutirostris" ~ "Cumacea"       ,  
Artal == "1999" & Flokkun == "Eudorella emarginata" ~ "Cumacea",
       TRUE ~ Flokkun)) %>% 
     drop_na() 

   bp <- boxplot.stats(lagfaerd_heiti$Nfm)
outliers <- bp$out   
lagfaerd_heiti <-  lagfaerd_heiti[lagfaerd_heiti$Nfm %in% outliers,]
   
   
# Define a function to find the taxon match based on the exact match and taxonomic hierarchy
find_taxon_match <- function(x, kolgr_taxa, y) {
  # Initialize a vector to store the matched taxon names
  matched_taxon <- rep(NA_character_, length(x))

  # Define the taxonomic hierarchy order
  taxon_levels <- c("Genus", "Family", "Superfamily", "Order", "Superorder", "Subterclass", "Infraclass", "Subclass", "Class", "Phylum", "Kingdom")

  # Iterate through each element in x (lagfaerd_heiti$Flokkun)
  for (i in 1:length(x)) {
    found_match <- FALSE
    
    # Try to find an exact match between the current element and the cleaned taxon names
    taxon_name <- find_exact_match(x[i], y$`Cleaned Taxon name`)
    if (!is.na(taxon_name)) {
      matched_taxon[i] <- taxon_name
      found_match <- TRUE
    }
    
    # If no exact match was found, search for a match in the taxonomic hierarchy
    if (!found_match) {
      for (level in taxon_levels) {
        taxon_name <- find_exact_match(kolgr_taxa[[level]][i], y$`Cleaned Taxon name`)
        if (!is.na(taxon_name)) {
          matched_taxon[i] <- taxon_name
          break
        }
      }
    }
  }

  # Return the vector of matched taxon names
  return(matched_taxon)
}

# Find the taxon match for each element in lagfaerd_heiti$Flokkun
lagfaerd_heiti$matched_taxon_name <- find_taxon_match(lagfaerd_heiti$Flokkun, lagfaerd_heiti, table_df)

# Identify rows with unmatched taxon names (empty strings)
unmatched_rows <-  lagfaerd_heiti$matched_taxon_name == ""

# Define a function to find the closest string match between two sets of strings
find_closest_match <- function(x, y) {
  string_distances <- stringdist::stringdistmatrix(tolower(x), tolower(y), method = "jw")
  y[apply(string_distances, 1, which.min)]
}

# For the unmatched rows, find the closest string match between Flokkun and Cleaned Taxon name
lagfaerd_heiti$matched_taxon_name[unmatched_rows] <- find_closest_match(lagfaerd_heiti$Flokkun[unmatched_rows], table_df$`Cleaned Taxon name`)

# Merge the data frames based on the matched taxon names
merged_data_kolgr <- merge(lagfaerd_heiti, table_df, by.x = "matched_taxon_name", by.y = "Cleaned Taxon name")


#write.csv(merged_data_kolgr,"docs/merged_data_alltB.csv", row.names = F)
```

```{r Feeding_Guilds}
merged_data_allt <- read.csv("docs/merged_data_allt.csv")
merged_data <- subset(merged_data_allt, stod %in% c("C4", "A7", "B5", "B8", "E4", "E3"))
colnames(merged_data) <- janitor::make_clean_names(colnames(merged_data))


data1_agg <- ddply(merged_data,.(artal, stod, feeding_guild),summarise,N=sum(nfm))
# Prepare data for ordination
data1_wide <- data1_agg %>% spread(key = feeding_guild, value = N, fill = 0)

#Fix column name for Artal to artal
colnames(kornast_organic_tap_dypi)[2]<-"artal"
merged_data <- merge(data1_wide, kornast_organic_tap_dypi, by = c("artal", "stod")) # Hér dettur 1999 út
rownames(merged_data) <- paste(merged_data$artal, merged_data$stod, sep = "_")
merged_data <- merged_data[, -c(1,2)] # Remove the 'Year' and 'Station' columns

# Separate environmental and species data
env_data <- merged_data[, c("20um","63um","125um","250um","1000um","tap","sd","dypi", "Síld 2013")]
guild_data <- merged_data[, !(names(merged_data) %in% c("20um","63um","125um","250um","1000um","tap","sd","dypi", "Síld 2013"))]

# Perform RDA or CCA
rda_result <- rda(guild_data ~ `20um` + `63um` + `125um` + `250um` + `1000um` + tap + sd + dypi + `Síld 2013`, data = env_data)
cca_result <- cca(guild_data ~ `20um` + `63um` + `125um` + `250um` + `1000um` + tap + sd + dypi + `Síld 2013`, data = env_data)

```



```{r env-group}

library(vegan)

# Define groups of environmental variables
group1 <- env_data[, c("20um", "63um", "125um", "250um", "1000um")] # Particle sizes
group2 <- env_data[, c("tap", "sd", "dypi", "Síld 2013")] # Other environmental factors

# Perform variation partitioning
vp_result <- varpart(guild_data, group1, group2)

# Print the results
print(vp_result)


```


The variation partitioning results show the unique and shared contributions of the two sets of environmental variables (group1: particle sizes, and group2: other environmental factors) to the feeding guild composition. The adjusted R-squared values indicate the proportion of variance in guild composition that can be explained by each set of variables or their shared effect.

Here's a summary of the results:

[a]: The unique contribution of particle sizes (group1) is 0.08331, which means that ~8.33% of the variance in guild composition can be explained by particle sizes alone.
[b]: The unique contribution of other environmental factors (group2) is 0.29846, which means that ~29.85% of the variance in guild composition can be explained by these factors alone.
[c]: The shared contribution of both sets of variables is 0.13334, which means that ~13.33% of the variance in guild composition can be explained by the combined effect of both sets of variables.
[d]: The unexplained variance (residuals) is 0.48490, which means that ~48.49% of the variance in guild composition cannot be explained by the environmental variables in your dataset.
Please note that collinearity was detected in group1 and in the combined set of variables (group1 + group2). This may affect the interpretation of the results, so it is essential to verify the assumptions of the RDA and check for multicollinearity among the variables. You can use the vif.cca() function from the 'vegan' package to compute variance inflation factors (VIF) and identify collinear variables. If necessary, consider removing or combining collinear variables to improve your analysis.

To test the significance of the individual fractions (a, b, and a+b+c), you can use the rda() function from the 'vegan' package, followed by the anova() function to perform a permutation test. Here's an example:


```{r capscale}




# Test the significance of the unique contribution of particle sizes (group1)
test_a <- rda(guild_data ~ `20um` + `63um` + `125um` + `250um` + `1000um`, data = env_data)
anova(test_a, by = "margin", permutations = 999)

# Test the significance of the unique contribution of other environmental factors (group2)
test_b <- rda(guild_data ~ tap + sd + dypi + `Síld 2013`, data = env_data)
anova(test_b, by = "margin", permutations = 999)

# Test the significance of the combined effect of both sets of variables (group1 + group2)
test_ab <- rda(guild_data ~ `20um` + `63um` + `125um` + `250um` + `1000um` + tap + sd + dypi + `Síld 2013`, data = env_data)
anova(test_ab, by = "margin", permutations = 999)



# Load the vegan package
library(vegan)

# Calculate variance inflation factors for each variable
vif_values <- caTools::vif(lm(`1000um` ~ `20um` + `63um` + `125um` + `250um`, data = particle_sizes))
print(vif_values)

# Check if any VIF values are above a threshold of 10
if (any(vif_values > 10)) {
  # Identify the index of the variable with the highest VIF
  index_to_remove <- which.max(vif_values)
  # Remove the variable with the highest VIF value and recalculate VIF values
  particle_sizes_vif_filtered <- particle_sizes[, -index_to_remove]
} else {
  particle_sizes_vif_filtered <- particle_sizes
}

# Rerun the analysis with the filtered dataset
test_a_filtered <- rda(guild_data ~ `20um` + `63um` + `125um` + `250um` + `1000um`, data = env_data)
anova(test_a_filtered, by = "margin", permutations = 999)


# Prepare the combined particle sizes and other environmental data
combined_data <- cbind(particle_sizes, env_data[, c("tap", "sd", "dypi", "Síld 2013")])

# Perform partial RDA using capscale
prda_result <- capscale(guild_data ~ . - Condition(`20um` + `63um` + `125um` + `250um` + `1000um`), data = combined_data)

# Run permutation tests for the partial RDA
anova(prda_result, by = "margin", permutations = 999)

```



